The Leaky Semicolon (Supplementary Material)

### 2021-11-10 paper/outro.tex
add Simon and Mark's funding acknowledgement
pass over intro
pass through 9-10

This paper is the first to present a direct denotational semantics for
sequential composition in a relaxed memory model that can be efficiently
compiled to modern CPUs. We extract from this model a semantic dependency
relation and use it to build \PwTc, a solution to the Out-of-Thin-Air problem
in \cXI, and \PwTmca{}, a model intended for safe languages such as Java and
Javascript.  Our work has several limitations, providing opportunities for
future work.

We have mechanized some proofs, but not all.  In particular, we have only a
pen-and-paper proof showing that \PwTmca{} supports optimal lowering to
\armeight{}.  The same is true for local data race-freedom (\ldrfsc{}).
Additionally, our proof sketch for \ldrfsc{} elides \RMW{}s, which have
caused complications in other models \cite{promising-ldrf}.

@@

\PwTmca{} does not validate access elimination: store-forwarding and
dead-write-removal are unsound.  We expect that these can be validated by
allowing events with different actions to merge. % in some instances.
Nor does \PwTmca{} validate the reverse inclusions for
\reflem{lem:if}\eqref{lem:ifelse:if:if}. The culprit is $\rdelay$, which
introduces order regardless of whether preconditions are disjoint.  As an
example, using augmentation, 
\begin{math}
  \sem{\IF{r}
  \THEN \PW{x}{1}
  \ELSE \PW{x}{2}
  \FI}
\end{math}
has an execution with
\begin{math}
  (r{=}0\mid\DW{x}{2})
  \xwki
  (r{\neq}0\mid\DW{x}{1}),
\end{math}
whereas
\begin{math}
  \sem{
    \IF{r} \THEN \PW{x}{1}\FI
    \SEMI
    \IF{\BANG r} \THEN \PW{x}{2}\FI
  \FI}
\end{math}
has no such execution.  We expect that this can be remedied by encoding
$\rdelay$ in the logic.
%\todo{Comment on \PwTpo{} and \PwTc{}?}

\PwTmca{1} is a simpler model than \PwTmca{2}, but requires fences on
acquiring reads for \armeight{}.  It would be illuminating to find out what
the performance penalty is for these fences.  In a similar vein, it would be
interesting to know if there is a performance penalty for banning
read-introduction, which is not valid for \PwT{}.

@@

Supplementary material for this paper available at
\url{https://weakmemory.github.io/pwt}.

\subsection*{Acknowledgements}
This paper has been greatly improved by the comments of the anonymous reviewers.

James Riely was supported by the National Science Foundation under
grant No.~CCR-1617175.

\todo{Add support for Mark\&Simon.}

Anton Podkopaev was supported by JetBrains Research.

>>>>

This paper is the first to present a direct denotational semantics for
sequential composition that can be efficiently compiled to modern \CPU{}s.  We
defined two models: \PwTc{} solves the out-of-thin-air problem for \cXI, and
\PwTmca{} solves it for safe languages such as Java and Javascript.

Our work has several limitations, providing opportunities for future work:

As mentioned in \textsection\ref{sec:related}, some safety guarantees may be
violated when read-introduction is followed by if-introduction, branching on
the read just introduced.  Nonetheless, read-introduction is ubiquitous in
some compilers \cite{DBLP:conf/pldi/LeeKSHDMRL17}.  It would be interesting
to know the cost of restricting this optimization.
%  
In a similar vein, \PwTmca{1} is a simpler model than \PwTmca{2}, but
requires fences on acquiring reads for \armeight{}.  It would be illuminating
to find out what the performance penalty is for these fences.

We have defined the soundness of compiler optimizations in the model, rather
than contextually: $\aCmd'$ is a sound refinement of $\aCmd$ if
$\sem{\aCmd'}\subseteq\sem{\aCmd}$.  This approach has several
advantages---for example, it is immediate that a sound optimization is sound
in any context.  It also has a disadvantage: some optimizations complicate
the semantics.  For example, \PwTmca{} does not validate access elimination,
such as store-forwarding and dead-write-removal---consider that complete
executions of
\begin{math}
  \sem{\PW{x}{1}\SEMI\PR{x}{r}}
\end{math}
must include a read action and that complete executions of
\begin{math}
  \sem{\PW{x}{1}\SEMI\PW{x}{2}}
\end{math}
must include two write actions.  As another example, \PwTmca{} does not
validate the reverse inclusions for
\reflem{lem:if}\eqref{lem:ifelse:if:if}---consider that
\begin{math}
  \sem{\IF{r}
  \THEN \PW{x}{1}
  \ELSE \PW{x}{2}
  \FI}
\end{math}
has an augmented (\reflem{lem:augment}) execution with
\begin{math}
  (r{=}0\mid\DW{x}{2})
  \xwki
  (r{\neq}0\mid\DW{x}{1}),
\end{math}
whereas
\begin{math}
  \sem{
    \IF{r} \THEN \PW{x}{1}\FI
    \SEMI
    \IF{\BANG r} \THEN \PW{x}{2}\FI
  \FI}
\end{math}
has no such execution.  We expect that these optimizations can be validated,
at the cost of complicating the semantics.  For access elimination, it is
likely sufficient to allow events with different actions to merge.  For
\reflem{lem:if}\eqref{lem:ifelse:if:if}, it is likely sufficient to encoding
$\rdelay$ in the logic---the problem in the execution above is that $\rdelay$
introduces order regardless of whether preconditions are disjoint.
%\todo{Comment on \PwTpo{} and \PwTc{}?}

@@

While we have mechanized some proofs, most of our proofs are informal.
In particular, we have only a pen-and-paper proof showing that \PwTmca{}
supports optimal lowering to \armeight{}.  The same is true for local data
race-freedom (\ldrfsc{})---additionally, our proof sketch for \ldrfsc{}
elides \RMW{}s, which have caused complications in other models
\cite{promising-ldrf}.

@@

Supplementary material for this paper is available at
\url{https://weakmemory.github.io/pwt}.

\subsection*{Acknowledgements}
This paper has been greatly improved by the comments of the anonymous reviewers.
%
James Riely was supported by the National Science Foundation under
grant No.~CCR-1617175.
%
Mark Batty and Simon Cooksey were supported by the EPSRC under grant
Nos.~EP/V000470/1 and~EP/R032971/1, and by VeTSS.
%
Anton Podkopaev was supported by JetBrains Research.

### 2021-11-10 paper/related.tex
accepted!
pass through 9-10

Other studies of relaxed memory can be categorized by their approach to
dependency calculation.  Hardware models use syntactic dependencies
\cite{alglave}.  Many software models do not bother with dependencies at all
\cite{Batty:2011:MCC:1926385.1926394, DBLP:journals/pacmpl/WattRP19,
  DBLP:conf/pldi/WattPPBDFPG20, goMM}.  Others have strong dependencies that
disallow compiler optimizations and efficient implementation, typically
requiring fences for every relaxed read on Arm
\cite{Lamport:1979:MMC:1311099.1311750, DBLP:conf/pldi/LahavVKHD17,
  Dolan:2018:BDR:3192366.3192421, DBLP:conf/lics/JeffreyR16,
  Boehm:2014:OGA:2618128.2618134}. %, DBLP:journals/corr/abs-1804-04214}.
%
Many of the most prominent models are operational models based on speculative
execution \cite{Manson:2005:JMM:1047659.1040336,
  DBLP:conf/esop/JagadeesanPR10,
  DBLP:conf/popl/KangHLVD17,DBLP:journals/pacmpl/ChakrabortyV19,DBLP:conf/pldi/LeeCPCHLV20,promising-ldrf}.

@@

% \todo{We agree that a new memory model needs to be positioned against existing
% models.  The usual result here is a compilation correctness to hardware
% memory models.  For PwT-MCA, we address this by showing compilation result
% for Armv8 model (§5).
% %
% Comparing software models, however, is unsatisfying: they are all
% incomparable, i.e., there are examples which are allowed by one/disallowed by
% the other and vice versa.
% %
% Morally, our model sits between the strong models (exemplified by RC11
% [Lahav-al:PLDI17]) and the speculative models (exemplified by the promising
% semantics [Kang-al:POPL17]).  As we argue in §1:
% %
% - The strong models require too much synchronization.
% %
% - The speculative models allow thin air behaviors.
% %
% Looking at the details, however, PwT-MCA is incomparable to both RC11 and
% promising semantics.  RC11 allows non-MCA behaviors that PwT-MCA disallows.
% PwT-MCA has a weaker notion of coherence than the promising semantics.
% %
% Some differences reflect an attempt to fix a bug.  For example, Weakestmo
% [Chakraborty-Vafeiadis:POPL19] purposefully disallows thin-air-like
% behaviors of the promising semantics.
% %
% Other differences reflect a different balance between allowed optimizations
% and reasoning principles.  There are fundamental conflicts, for example:
% %
% - between Common Subexpression Elimination (CSE) and read-read coherence
%   (§D.1)
% %
% - between if-introduction (aka, case analysis, if-closure) and java-style
%   final-field semantics.  (If-introduction requires that address dependencies
%   and control dependencies are the same.  Final-fields require that they be
%   different.)
% }
\citet{DBLP:journals/pacmpl/JagadeesanJR20} argue that the speculative models
allow too many executions, resulting in a failure of temporal reasoning and
potentially jeopardizing type safety and other security properties.  In a
similar vein, \citet{promising-ldrf} argue that local \drf{} guarantees are
violated when read-introduction is followed by if-introduction, branching on
the read just introduced.  These optimizations are validated by the
speculative models---\citeauthor{promising-ldrf} manage to avoid the problem
by adopting a sub-optimal lowering for \RMW{}s.  \PwT{} does not suffer from
this problem, since \PwT{} does not validate read-introduction.  Nonetheless,
read-introduction is ubiquitous in some compilers
\cite{DBLP:conf/pldi/LeeKSHDMRL17}.  There appears to be a genuine tension
between temporal reasoning, as supported by \PwT{}, and read-introduction, as
supported by the speculative models.

Other work in relaxed memory has shown that tooling is especially useful to
researchers, architects, and language specifiers, enabling them to build
intuitions
experimentally~\cite{DBLP:conf/esop/PaviottiCPWOB20,PrideMM,alglave,Batty:2011:MCC:1926385.1926394}.
Unfortunately, it is not obvious that tools can be built for all
thin-air-free models, the calculation of
\citet{Pichon-Pharabod:2016:CSR:2837614.2837616} does not have a termination
proof for an arbitrary input, and the enormous state space for the
operational models of \citet{DBLP:conf/popl/KangHLVD17} and
\citet{DBLP:journals/pacmpl/ChakrabortyV19} is a daunting prospect for a tool
builder -- and as yet no tool exists for automatically evaluating these
models.  We described a tool, \PwTer, for automatically evaluating \PwT{} in
\textsection\ref{sec:tool}.

>>>>

Other studies of relaxed memory can be categorized by their approach to
dependency calculation.  Hardware models use syntactic dependencies
\cite{alglave}.  Many software models do not bother with dependencies at all
\cite{Batty:2011:MCC:1926385.1926394, DBLP:journals/pacmpl/WattRP19,
  DBLP:conf/pldi/WattPPBDFPG20, goMM}.  Others have strong dependencies that
disallow compiler optimizations and efficient implementation, typically
requiring fences for every relaxed read on \armeight{}
\cite{Lamport:1979:MMC:1311099.1311750, DBLP:conf/pldi/LahavVKHD17,
  Dolan:2018:BDR:3192366.3192421, DBLP:conf/lics/JeffreyR16,
  Boehm:2014:OGA:2618128.2618134}. %, DBLP:journals/corr/abs-1804-04214}.
%
Many of the most prominent models are operational models based on speculative
execution \cite{Manson:2005:JMM:1047659.1040336,
  DBLP:conf/esop/JagadeesanPR10,
  DBLP:conf/popl/KangHLVD17,DBLP:journals/pacmpl/ChakrabortyV19,DBLP:conf/pldi/LeeCPCHLV20,promising-ldrf}.

@@

% \todo{We agree that a new memory model needs to be positioned against existing
% models.  The usual result here is a compilation correctness to hardware
% memory models.  For PwT-MCA, we address this by showing compilation result
% for Armv8 model (§5).
% %
% Comparing software models, however, is unsatisfying: they are all
% incomparable, i.e., there are examples which are allowed by one/disallowed by
% the other and vice versa.
% %
% Morally, our model sits between the strong models (exemplified by RC11
% [Lahav-al:PLDI17]) and the speculative models (exemplified by the promising
% semantics [Kang-al:POPL17]).  As we argue in §1:
% %
% - The strong models require too much synchronization.
% %
% - The speculative models allow thin-air behaviors.
% %
% Looking at the details, however, PwT-MCA is incomparable to both RC11 and
% promising semantics.  RC11 allows non-MCA behaviors that PwT-MCA disallows.
% PwT-MCA has a weaker notion of coherence than the promising semantics.
% %
% Some differences reflect an attempt to fix a bug.  For example, Weakestmo
% [Chakraborty-Vafeiadis:POPL19] purposefully disallows thin-air-like
% behaviors of the promising semantics.
% %
% Other differences reflect a different balance between allowed optimizations
% and reasoning principles.  There are fundamental conflicts, for example:
% %
% - between Common Subexpression Elimination (CSE) and read-read coherence
%   (§D.1)
% %
% - between if-introduction (aka, case analysis, if-closure) and java-style
%   final-field semantics.  (If-introduction requires that address dependencies
%   and control dependencies are the same.  Final-fields require that they be
%   different.)
% }
\citet{DBLP:journals/pacmpl/JagadeesanJR20} argue that the speculative models
allow too many executions, resulting in a failure of temporal reasoning and
potentially jeopardizing type safety and other security properties.  In a
similar vein, \citet{promising-ldrf} argue that local \drf{} guarantees are
violated when read-introduction is followed by if-introduction, branching on
the read just introduced.  These optimizations are validated by the
speculative models---\citeauthor{promising-ldrf} manage to avoid the problem
by adopting a sub-optimal lowering for \RMW{}s.  \PwT{} does not suffer from
this problem, since \PwT{} does not validate read-introduction.  There appears to be a genuine tension
between temporal reasoning, as supported by \PwT{}, and read-introduction, as
supported by the speculative models.

Other work in relaxed memory has shown that tooling is especially useful to
researchers, architects, and language specifiers, enabling them to build
intuitions
experimentally~\cite{DBLP:conf/esop/PaviottiCPWOB20,PrideMM,alglave,Batty:2011:MCC:1926385.1926394}.
Unfortunately, it is not obvious that tools can be built for all
thin-air-free models: the calculation of
\citet{Pichon-Pharabod:2016:CSR:2837614.2837616} does not have a termination
proof for an arbitrary input; the enormous state space for the
operational models of \citet{DBLP:conf/popl/KangHLVD17} and
\citet{DBLP:journals/pacmpl/ChakrabortyV19} is daunting for a tool
builder---and as yet no tool exists for automatically evaluating these
models.  We described a tool for automatically evaluating \PwT{} in
\textsection\ref{sec:tool}.
